Are environmental and geographic variables effective surrogates for genetic variation in conservation planning?

### 2016-02-05 article/article.Rmd
+ replace adegenet with mclust

Protected areas are one of the hallmarks of modern conservation [@r440]. They buffer species from anthropogenic impacts and provide places for biodiversity processes to persist (eg. nutrient cycling and water purification). Additionally, protected areas set the stage for direct management actions to enhance species recovery (eg. habitat rehabilitation, translocation, and invasive species management). Since the resources available for conservation are limited, protected areas must be cost-effective. As a consequence, plans for protected areas (prioritizations) are often generated using decision support tools [eg. $\texttt{MARXAN}$; @r22; $\texttt{Zonation}$; @r429]. These tools often formulate the benefit that a given species receives from a solution based on the proportion of its range that is protected. However, the true benefit that a species might receive is dependent on many other additional factors.

Protected areas must preserve intra-specific genetic variation to maximize species recovery [@r67; @r3; @r61]. This variation can be classified as either adaptive or neutral [reviewed in @r????]. Adaptive genetic variation is associated with loci currently under selection, whereas neutral genetic variation is associated with loci that are not. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral variation reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze, and as a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus many prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver conservation outcomes.

Recently, conservation scientists have begun investigating surrogates to generate more effective prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is driven by selection pressures, environmental variables have been explored as potential surrogates for genetic variation [eg. @r247; @r2]. On the other hand, neutral variation arises from a reduction in gene flow, and so variables that predict reduction in gene flow between areas have been explored. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have lower gene flow, and in turn have greater neutral variation among them. The use of distance-based surrogates has been found to improve prioritizations in an island network [@r200]. However the effectiveness of distance-based surrogates remains untested in contiguous systems conventionally used in conservation planning where gene flow is complicated by additional factors (eg. elevation in terrestrial systems). Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, if these surrogates are ineffective, conservation planners may waste precious resources on places with little benefit.

Our overarching aim was to investigate the effectiveness of environmental and geographic surrogates for adaptive neutral genetic variation in the context of conservation planning. First, we aimed to understand the relationship between the surrogate targets used to generate a prioritization and the proportion of genetic variation secured in the prioritisation. Second, we aimed to determine if the addition of surrogate-based targets in conservation planning exercises would result in prioritizations that secured substantially more genetic variation. If these surrogatess can be shown to improve the proportion of genetic variation secured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

## Methods
### Study system
To address the aims of the study, we utilized species distribution and genomic data collected by the IntraBioDiv project [@r451; see @r452 for methodological details]. This dataset was particularly well suited; it provides spatially explicit genomic data for a multitude of species across a broad geographic area over large environmental gradients. As part of this project, data were collected for 27 alpine plant species were across the European Alps according to a grid (22.3 km $\times$ 25 km; Figure S1). Briefly, in every second cell, plant samples were collected from three individuals for each species present inside the cell. Cells that did not contain any plant samples were omitted from our analysis. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453]. Matrices denoting the presence/absence of polymorphisms at loci were constructed independently for each species. Thus for each second grid cell, for each of the species in the grid cell, the dataset contains information describing the genomic variation of individuals inside the grid cell. We used these grid cells as planning units to develop prioritizations to investigate the effectiveness of surrogates for the (AFLP) genetic variation. Most of our analyses described hereafter were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Landscape data
We aimed to investigate the effectiveness of climatic and geographic surrogates. To describe the location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$), calculated their centroids, and extracted its two-dimensional coordinates. To describe the climatic variation among each cell, we obtained 19 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principle components analysis (PCA; Table S1). The first two principle components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2). To examine if the influence of opportunity cost, we calculated the total human population density inside each planning unit (obtained at 1 km resolution from the Global Rural-Urban Mapping Project; GRUMP V1; @r472).

### Outlier locus detection
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first need to identify which of the sampled loci--if any--were under selection for each species. We used outlier locus detection methods [evaluated in @r454] to achieve this. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. The advantage of these methods, in contrast with environmental association analyses [eg. r@XXX], is that they do not use environmental data to identify loci under selection, which would have biased our analysis. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this dataset (_Getiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the dataset.

First, we grouped individuals into genetic lineages for each species. The purpose of this step was to avoid misclassifying loci as adaptive. We used discriminant analyses of principle components [DAPC; @r467] to assign individuals to populations [implemented in $\texttt{adegenet}$ R package; @r468]. Briefly, for each species, we transformed the binary loci data using a principle components analysis (PCA) and obtained the principle components required to explain at least `r adegenet.params.LST[[MODE]]$perc.pca` % variation. We subject these components to multiple \textit{k}-means clustering models with varying numbers of assumed populations (\textit{k} = 1--`r adegenet.params.LST[[MODE]]$max.n.clust`; `r adegenet.params.LST[[MODE]]$n.iter` iterations; `r adegenet.params.LST[[MODE]]$n.start` random starts). We then  applied Ward's clustering method to the Bayesian Information Criterion (BIC) statistics associated with each model to partition the models into two groups, and selected the model before the first group switch (as recommended in the \texttt{adegenet} manual; Figure S4). The populations memberships associated with these models were used for subsequent analysis (Figure S5).

We identified loci under selection for each species (Table S2). For each species, we ran $\texttt{Bayescan}$ [version 2.1; @r456] using the population memberships identified in the previous analysis (1:1 prior odds; `r bayescan.params.LST[[MODE]]$nbp` pilot runs; `r bayescan.params.LST[[MODE]]$burn` burn-in iterations; `r bayescan.params.LST[[MODE]]$n` post-burn-in iterations thinned by `r bayescan.params.LST[[MODE]]$thin` iterations). We omitted loci where the global frequency of the minor allele was $\geq 0.05$ to avoid false-positives (following guidelines in the $\texttt{Bayescan}$ manual). To ensure convergence, we ran `r bayescan.params.LST[[MODE]]$reps` replicates per species using a suitable false discovery rate [FDR; $\leq `r bayescan.params.LST[[MODE]]$fdr`$; @r466].

After classifying loci as adaptive or neutral, we mapped the main gradients of the adaptive (if detected) and neutral genetic variation for each species (Figures S6--S32). For each species, for each type of genetic variation detected, we used Gower distances [@r464] to express the genomic differences between individuals [using the $\texttt{cluster}$ R package to accommodate sparsity; @r457]. These distances were subject to multiple non-metric multi-dimensional scaling analyses [NMDS; implemented in the $\texttt{vegan}$ R package; @r458] using the minimum number of dimensions required to achieve an adequate stress value  ($\leq$ `r mds.params.LST[[MODE]]$max.stress`; `r mds.params.LST[[MODE]]$trymax` random starts for all models; Table S3). Since each grid cell had up to three samples per species, we used the average of the ordinated values associated with the samples in each grid cell to characterize their the genetic variation of individuals in the cells.

@@

### Limitations/Future directions
We wish to alert practitioners to several limitations associated with our analysis. First, the size of the planning units we used (23 \times 25 km) is much larger than typically used in conservation planning scenarios (approx. 1--5 km^2). We used this resolution because the genetic data was collected at this scale. Whilst we could have interpolated the genetic data to a finer resolution and used smaller planning units, this would have biased our analysis because the surrogate variables are spatially auto-correlated. Second, we used geographic distances as surrogates for connectivity between planning units. Although connectivity between areas can be modeled using isolation by resistance measures derived from landscape data [eg. elevation; reviewed in @rXX], they are unlikely to be of use to conservation planners since they require genetic data to be tuned. Third, we used AFLP data to characterize the main gradients of genetic variation. Whilst next-generation sequencing provides higher resolution genetic information than AFLP data [reviewed in @rXXX], our methodology would still have utilised only the main gradients of the genetic variation to generate prioritizations in a feasible amount of time. Overall, the limitations associated with this study are relatively minor.

>>>>

Protected areas are one of the hallmarks of modern conservation [@r440]. They buffer species from anthropogenic impacts and provide places for biodiversity processes to persist (eg. nutrient cycling and water purification). Protected areas also set the stage for direct management actions to enhance species recovery (eg. habitat rehabilitation, translocation, and invasive species management). But the resources available for conservation are limited, and so protected areas must be cost-effective. As a consequence, decision support tools [eg. $\texttt{MARXAN}$; @r22; $\texttt{Zonation}$; @r429] are used to identify cost-effective plans for protected areas (prioritizations). These tools typically estimate the benefit that a given species receives from a given prioritization based on the proportion of its range that is secured (eg. using targets). However, the true benefit of a given prioritization is dependent on many additional factors that are not used to inform the reserve selection process.

Protected areas must preserve intra-specific genetic variation to maximize species recovery [@r67; @r3; @r61]. This variation can be classified as either adaptive or neutral [reviewed in @r????]. Adaptive genetic variation is associated with loci currently under selection. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that are not under natural selection. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze. As a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus many prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver conservation outcomes.

In light of this problem, conservation scientists have been investigating surrogates for genetic variation to generate better prioritizations--but without the overheads associated with genetic data. Since adaptive genetic variation is ultimately driven by selection pressures, conservation planners have used the distribution of likely selection pressures. For instance, environmental variation such as local habitat type [@r247] and climate [@r2] have been used to guide the reserve selection process. While such variables have been used to detect and verify adaptive genetic variation [eg. @r475; @r455], the effectiveness of prioritizations based on such surrogates remains unverified.

On the other hand, neutral genetic variation arises a reduction in gene flow between individuals. By modeling the connectivity between areas, a function of gene flow, conservation scientists have explored the effectiveness of surrogates for neutral genetic variation. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have reduced connectivity, and in turn exchange less genetic material. The use of geographic distance-based surrogates has been found to improve prioritizations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain; eg. @r474]--the systems commonly featured in real-world conservation planning exercises. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, if these surrogates are ineffective, conservation planners may waste precious resources on places with little benefit.

Our overarching aim was to investigate the effectiveness of environmental and geographic surrogates for adaptive neutral genetic variation in the context of conservation planning. First, we aimed to understand the relationship between the surrogate targets used to generate a prioritization and the proportion of genetic variation secured in the prioritization. Second, we aimed to determine if the addition of surrogate-based targets in more realistic conservation planning scenarios would result in prioritizations that secured substantially more genetic variation. If these surrogates can be shown to improve the intra-specific genetic variation captured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

## Methods
### Study system
To address the aims of the study, we utilized species distribution and genomic data collected by the IntraBioDiv project [@r451; see @r452 for methodological details]. This dataset was particularly well suited; it provides spatially explicit genomic data for a multitude of species across a broad geographic area over large environmental gradients. As part of this project, data were collected for 27 alpine plant species were across the European Alps according to a grid (22.3 km $\times$ 25 km; Figure S1). Briefly, in every second cell, plant samples were collected from three individuals for each species present inside the cell. Cells that did not contain any plant samples were omitted from our analysis. Samples were genotyped using amplified fragment length polymorphisms [AFLP; @r453]. Matrices denoting the presence/absence of polymorphisms at loci were constructed independently for each species. Thus for each second grid cell, for each of the species in the grid cell, the dataset contains information describing the genomic variation of individuals inside the grid cell. We used these grid cells as planning units to develop prioritizations to investigate the effectiveness of surrogates for the (AFLP) genetic variation. The majority of our analyses described hereafter were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Landscape data
We aimed to investigate the effectiveness of climatic and geographic surrogates. To describe the location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$), calculated their centroids, and extracted its two-dimensional coordinates. To describe the climatic variation among each cell, we obtained 19 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principle components analysis (PCA; using ArcMap version 10.2.2; Table S1). The first two principle components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2). To examine if the influence of opportunity cost, we calculated the total human population density inside each planning unit (obtained at 1 km resolution from the Global Rural-Urban Mapping Project; GRUMP V1; @r472).

### Outlier locus detection
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first need to identify which of the sampled loci--if any--were under selection for each species. We used outlier locus detection methods [evaluated in @r454] to achieve this. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. The advantage of these methods--in contrast with environmental association analyses [eg. r@475]--is that they do not use environmental data to identify loci under selection, which would have biased our analysis. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this dataset (_Getiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the dataset.

First, we grouped individuals into genetic lineages for each species. The purpose of this step was to avoid misclassifying loci as adaptive. We used Gaussian mixture models to assign individuals to populations [implemented in $\texttt{mclust}$ R package; @r477; @r467]. Briefly, for each species, we fit a suite of models to estimate the probability that individuals belonged different populations (using default priors and covariance structures). For each parametrization, we fit a set of models assuming varying numbers of total populations (\textit{G} = 1--`max(r mclust.params.LST[[MODE]]$G)`). We identified the most parsimonious models (lowest Bayesian Information Criterion $BIC$ statistic; Figures S4), and used these to group individuals into populations (Figure S5).

We identified loci under selection. For each species, we ran $\texttt{Bayescan}$ [version 2.1; @r456] using the population memberships identified in the previous analysis (1:1 prior odds; `r bayescan.params.LST[[MODE]]$nbp` pilot runs; `r bayescan.params.LST[[MODE]]$burn` burn-in iterations; `r bayescan.params.LST[[MODE]]$n` post-burn-in iterations thinned by `r bayescan.params.LST[[MODE]]$thin` iterations; Table S2). To avoid false-positive, we omitted loci where the global frequency of the minor allele was $\geq 0.05$ (following guidelines in the $\texttt{Bayescan}$ manual). Additionally, to prevent individuals with unclear ancestry from biasing this analysis, we omitted individuals if there was considerable uncertainty in their population membership (maximum probability $\leq `r mclust.params.LST[[MODE]]$probability.threshold`$ % ). We ran `r bayescan.params.LST[[MODE]]$reps` replicates per species using a suitable false discovery rate to ensure convergence [FDR $\leq `r bayescan.params.LST[[MODE]]$fdr`$; @r466].

After classifying loci as adaptive or neutral, we mapped the main gradients of the adaptive (if detected) and neutral genetic variation for each species (Figures S6--S32). For each species, for each type of genetic variation detected, we used Gower distances [@r464] to express the genomic differences between individuals [using the $\texttt{cluster}$ R package to accommodate sparsity; @r457]. These distances were subject to multiple non-metric multi-dimensional scaling analyses [NMDS; implemented in the $\texttt{vegan}$ R package; @r458] using the minimum number of dimensions required to achieve an adequate stress value  ($\leq$ `r nmds.params.LST[[MODE]]$max.stress`; `r nmds.params.LST[[MODE]]$trymax` random starts for all models; Table S3). Since each grid cell had up to three samples per species, we used the average of the ordinated values associated with the samples in each grid cell to summarize the typical genetic characteristics of individuals in the cells.

@@

### Limitations/Future directions
We wish to alert practitioners to several limitations associated with our analysis. First, the size of the planning units we used (23 \times 25 km) is much larger than typically used in conservation planning scenarios (approx. 1--5 km^2). We used this resolution because the genetic data was collected at this scale. Whilst we could have interpolated the genetic data to a finer resolution and used smaller planning units, this would have biased our analysis because the surrogate variables are spatially auto-correlated. Second, we used geographic distances as surrogates for connectivity between planning units. Although connectivity between areas can be modeled using isolation by resistance measures [@r473] derived from landscape data [eg. @r474], they are unlikely to be of use in broad-scale conservation exercises since they need to be tuned using species-specific parameters. Third, we used AFLP data to characterize the main gradients of genetic variation. Whilst next-generation sequencing provides higher resolution genetic information than AFLP data [reviewed in @rXXX], our methodology would still have utilized only the main gradients of the genetic variation to generate prioritizations in a feasible amount of time. Overall, the limitations associated with this study are relatively minor.
